---
title: "What Factors Influence College Graduate Earnings?"
author: "Aiden Akbarov"
date: "May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

# Overview

This analysis investigates the relationship between academic major characteristics and median earnings of recent college graduates in the United States. Using a dataset of 172 majors, I build a multiple linear regression model to identify the strongest predictors of post-graduation income.

**Research Questions:**

- Do majors with higher female representation earn less?
- Does unemployment rate or job type predict earnings?
- Which broad field of study is the strongest predictor?

---

# Setup

```{r libraries}
library(car)
library(ggplot2)
library(olsrr)      # Model selection and diagnostics
library(Metrics)    # MAE and MSE
library(rsample)    # Data partitioning
```

---

# Data Preparation

```{r data-prep}
df <- read.csv("../data/recent-grads.csv")
df <- df[complete.cases(df), ]   # Remove rows with missing values (Food Science major)

cat("Dataset dimensions:", nrow(df), "rows,", ncol(df), "columns\n")
cat("Response variable summary:\n")
summary(df$Median)
```

One observation (Food Science) was removed due to missing values in key variables including Total, Men, Women, and ShareWomen.

---

# Exploratory Data Analysis

## Distribution of Median Earnings

```{r eda-distribution, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

# Histogram
hist(df$Median,
     main = "Histogram of Median Earnings",
     xlab = "Median Earnings ($)",
     col  = "lightgray",
     breaks = 20)

# Boxplot
boxplot(df$Median,
        main = "Distribution of Median Earnings",
        ylab = "Median Earnings ($)",
        col  = "lightgray")

par(mfrow = c(1, 1))
```

The distribution is right-skewed — most majors earn between \$30,000 and \$45,000, with a smaller number earning significantly more. High-value outliers are primarily Engineering majors and were retained as they represent legitimate salary variation.

## Earnings by Major Category

```{r eda-category, fig.width=10, fig.height=7}
ggplot(df, aes(x = reorder(Major_category, Median, median), y = Median)) +
  geom_boxplot(fill = "lightgray") +
  coord_flip() +
  labs(title = "Median Earnings by Major Category",
       x = "Major Category",
       y = "Median Earnings ($)") +
  theme_minimal()
```

Engineering has the highest median earnings. Psychology & Social Work, Arts, and Humanities show the lowest. Variance within Business is notably high.

## Relationships with Predictor Variables

```{r eda-scatter, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))

plot(df$Unemployment_rate, df$Median,
     main = "Median Earnings vs Unemployment Rate",
     xlab = "Unemployment Rate", ylab = "Median Earnings ($)", pch = 1)

plot(df$ShareWomen, df$Median,
     main = "Median Earnings vs Share of Women",
     xlab = "Share of Women", ylab = "Median Earnings ($)", pch = 1)

plot(df$Low_wage_jobs, df$Median,
     main = "Median Earnings vs Low Wage Jobs",
     xlab = "Low Wage Jobs", ylab = "Median Earnings ($)", pch = 1)

plot(df$College_jobs, df$Median,
     main = "Median Earnings vs College Jobs",
     xlab = "College Jobs", ylab = "Median Earnings ($)", pch = 1)

par(mfrow = c(1, 1))
```

## Correlation and Collinearity Check

```{r eda-correlation}
num_vars <- df[, c("Median", "ShareWomen", "Unemployment_rate",
                   "College_jobs", "Non_college_jobs", "Low_wage_jobs")]
round(cor(num_vars, use = "complete.obs"), 4)
```

Key finding: `Non_college_jobs` and `Low_wage_jobs` are very highly correlated (r = 0.98). Including both would cause multicollinearity — this informed our model selection approach.

---

# Model Building

## Train / Test Split (70/30)

```{r split}
set.seed(4164)
data_split <- initial_split(df, prop = 0.70)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Ensure all Major_category levels in test set exist in training set
test_data <- test_data[test_data$Major_category %in% unique(train_data$Major_category), ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size: ", nrow(test_data), "\n")
```

## Full Model

```{r full-model}
full_model <- lm(Median ~ ShareWomen + Unemployment_rate +
                   College_jobs + Low_wage_jobs + Major_category,
                 data = train_data)
```

## Backward Elimination via AIC

```{r model-selection}
selection_results <- ols_step_backward_aic(full_model)
final_model       <- selection_results$model
summary(final_model)
```

Variables eliminated: `Unemployment_rate` and `College_jobs`. Final model retains `ShareWomen` and `Major_category` as the primary predictors.

---

# Model Diagnostics

## Multicollinearity — VIF

```{r vif}
ols_coll_diag(final_model)
```

All VIF values are below 4.0, confirming no significant multicollinearity in the final model.

## Homoscedasticity — Residuals vs Fitted

```{r residuals, fig.width=8, fig.height=5}
plot(final_model$fitted.values, residuals(final_model),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch  = 20, col = "darkblue")
abline(h = 0, col = "red", lty = 2)
```

Residuals are roughly randomly scattered around zero. A slight increase in spread at higher fitted values is driven by high-earning Engineering outliers — mild heteroscedasticity common in salary data.

## Normality of Residuals — Shapiro-Wilk Test

```{r normality}
cat("--- Original Model ---\n")
shapiro.test(residuals(final_model))
```

The original model shows departure from normality (p < 0.05). A log transformation was applied to investigate:

```{r log-transform}
train_data$log_Median <- log(train_data$Median)
test_data$log_Median  <- log(test_data$Median)

log_model <- lm(log_Median ~ ShareWomen + Major_category, data = train_data)

cat("--- Log-Transformed Model ---\n")
shapiro.test(residuals(log_model))
```

```{r log-residuals, fig.width=8, fig.height=5}
plot(log_model$fitted.values, residuals(log_model),
     main = "Residuals vs Fitted (Log-Transformed Model)",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch  = 20, col = "darkgreen")
abline(h = 0, col = "red", lty = 2)
```

The log-transformed model substantially improved normality (W = 0.972). However, the original-scale model was retained to preserve direct dollar-amount interpretability of coefficients. Both models yielded consistent predictor significance.

## Influential Observations — Cook's Distance

```{r cooks, fig.width=10, fig.height=5}
ols_plot_cooksd_bar(final_model)
```

Four observations (43, 79, 103, 120) exceeded the Cook's Distance threshold of 0.033. These were retained in the final model as they represent legitimate high-earning Engineering majors rather than data entry errors — removing them would reduce the representativeness of the dataset.

---

# Model Validation

```{r validation}
# Generate predictions on test set
test_preds <- predict(final_model, newdata = test_data)

# Performance metrics
test_mae <- mae(test_data$Median, test_preds)
test_mse <- mse(test_data$Median, test_preds)
null_mae  <- mae(test_data$Median, rep(mean(train_data$Median), nrow(test_data)))

cat("--- Validation Results ---\n")
cat("Testing MAE:          $", round(test_mae, 2), "\n")
cat("Baseline (Mean) MAE:  $", round(null_mae, 2), "\n")
cat("Improvement:          ", round((1 - test_mae/null_mae)*100, 1), "%\n")
```

```{r pred-plot, fig.width=8, fig.height=6}
ggplot(test_data, aes(x = Median, y = test_preds)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title    = "Model Accuracy: Predicted vs. Actual Earnings",
       subtitle = paste("Testing MAE: $", round(test_mae, 2)),
       x        = "Actual Median Earnings ($)",
       y        = "Predicted Median Earnings ($)") +
  theme_minimal()
```

---

# Results Summary

**Final Model:**

$$\text{Median} = 41{,}843 - 17{,}040(\text{ShareWomen}) + \beta_{\text{Major\_category}} + \varepsilon$$

**Key findings:**

- Engineering graduates earn ~\$20,400 more than the Agriculture baseline
- For every 10% increase in female representation, predicted median salary drops ~\$1,700
- The model explains **55% of variance** in median earnings (Adjusted R² = 0.545)
- Testing MAE of \$4,427 — roughly **47% better than the baseline**

---

# References

- Jun, E. (n.d.). *data_college_income* [Data set]. GitHub. https://github.com/emjun/data_college_income
- R Core Team. (2024). *R: A language and environment for statistical computing.* https://www.R-project.org/
- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis.* Springer-Verlag.
- Hebbali, A. (2020). *olsrr: Tools for Building OLS Regression Models.* https://cran.r-project.org/package=olsrr
